### 2022-10-10 Proposal/proposal.tex
added in all the words

\documentclass[12pt]{article}

%% preamble: Keep it clean; only include those you need

\usepackage[margin = 1in]{geometry}

% highlighting hyper links
\usepackage[colorlinks=true, citecolor=blue]{hyperref}

%% meta data

\title{Proposal}
\author{Olivia Frillici\\
  Department of Statistics\\
  University of Connecticut
}

\begin{document}
\maketitle
The topic of my final research statistical paper will be predicting heart disease. 17.9 million people die of cardiovascular disease globally each year(WHO). By researching and developing tools in which medical providers can input patient data and get outputs pertaining to if a patient has heart disease, medical providers can prescribe lifestyle changes to mitigate and decrease risk of significant cardiac events. 
The primary goal is to develop a model for predicting whether a person has heart disease or not. A secondary goal is to also explore some of the individual variables to get a better picture of the data. These goals are important because they possibly have real world ramifications, that could if proven accurate and valid change part of the medical landscape. 
The data is acquired from kaggle and is the “largest heart disease data set available”. The creator of the data set combined a total of 5 datasets which had 11 overlapping variables. There are 918 observations. From the kaggle page which holds the data the variables are as follows: 
Age: age of the patient [years]
Sex: sex of the patient [M: Male, F: Female]
ChestPainType: chest pain type [TA: Typical Angina, ATA: Atypical Angina, NAP: Non-Anginal Pain, ASY: Asymptomatic]
RestingBP: resting blood pressure [mm Hg]
Cholesterol: serum cholesterol [mm/dl]
FastingBS: fasting blood sugar [1: if FastingBS > 120 mg/dl, 0: otherwise]
RestingECG: resting electrocardiogram results [Normal: Normal, ST: having ST-T wave abnormality (T wave inversions and/or ST elevation or depression of > 0.05 mV), LVH: showing probable or definite left ventricular hypertrophy by Estes' criteria]
MaxHR: maximum heart rate achieved [Numeric value between 60 and 202]
ExerciseAngina: exercise-induced angina [Y: Yes, N: No]
Oldpeak: oldpeak = ST [Numeric value measured in depression]
ST_Slope: the slope of the peak exercise ST segment [Up: upsloping, Flat: flat, Down: downsloping]
HeartDisease: output class [1: heart disease, 0: Normal]
My statistical analysis plan is to use a classical decision tree to create a model which predicts if heart disease is present. I plan to separate the total number of observations into a training and validation set. After creating my initial model I plan to prune the classic tree because it is based on the training set and I want to attempt to mitigate the risks of overfitting the model onto the training set. After I feel that I have created a model to the best of my ability I will use the validation set to see how good my model is. I plan to assess the false positive and false negative rate along with the accuracy. Additionally if time allows I may attempt to find standards for supervised machine learning machines in medicine and compare my results to the standards to see if it could be usable or valid in the actual real world. In addition to creating a machine learning model I also plan to visualize some of the variables in a way which is easy to communicate and digest for the masses. 
I expect to find a model which can predict heart disease because this is the aim of the classical decision tree which I am creating. I also expect to find through data visualization some interesting sample proportions of certain categorical type variables. I expect this because inherently we are all individuals and so to see what the proportional breakdown i.e. the prevalence of some of the characteristics definitely peaks my curiosity.  Additionally I am curious to see if there are differences in box and whisker plots of more continuous variables that exist on a numerical scale between heart disease and non heart disease individuals. Potential impacts of this data analysis include possibly a functioning machine learning model that may help diagnose heart disease. If the investigation doesn’t work out in the way I am currently predicting or the data analysis predictions which I set above turn out to be wrong I imagine I will gain new knowledge. Sometimes failure can be the best teacher in the long run. Even if it goes wrong - I hope I can identify and learn from the mistakes and missteps I take so that I can grow and change because of this experience.
In conclusion the aim of this paper is to create a supervised machine learning model, a classical decision tree, to predict whether an individual has heart disease or not. The implications of the creation of an accurate model could have a significant impact not only on the medical community but also the public at large in an effort to increase their present and future health.

https://www.who.int/health-topics/cardiovascular-diseases#tab=tab_1
data kaggle citation:
fedesoriano. (September 2021). Heart Failure Prediction Dataset. Retrieved [Date Retrieved] from https://www.kaggle.com/fedesoriano/heart-failure-prediction.
\end {document}

### 2022-11-10 Manuscript/paper.tex
added outline structure

this will be my paper
\documentclass[12pt]{article}

%% preamble: Keep it clean; only include those you need
\usepackage{amsmath}
\usepackage[margin = 1in]{geometry}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{natbib}

% for space filling
\usepackage{lipsum}
% highlighting hyper links
\usepackage[colorlinks=true, citecolor=blue]{hyperref}

%% meta data

\title{Stats Paper}
\author{Olivia Frillici\\
  Department of Statistics\\
  University of Connecticut
}

\begin{document}
\maketitle
\begin{abstract}
    Here is the abstract.  
    \end{abstract}

    \section{Introduction}
    \label{sec:intro}

    Use this section to answer three questions:
    Why is the topic important/interesting?
    What has been done on this topic in the literature?
    What is your contribution?

    % roadmap
The rest of the paper is organized as follows.
The data will be presented in Section~\ref{sec:data}.
The methods are described in Section~\ref{sec:meth}.
The results are reported in Section~\ref{sec:resu}.
A discussion concludes in Section~\ref{sec:disc}.

\section{Data}
\label{sec:data}

Use this section to describe the data that helps to answer your research
questions.

\section{Methods}
\label{sec:meth}

Use this section to present the methodologies that will generate results by
analyzing the data.

\section{Results}
\label{sec:resu}

\section{Discussion}
\label{sec:disc}

What are the main contributions again?

What are the limitations of this study?

What are worth pursuing further in the future?

\bibliography{refs}
\bibliographystyle{mcap}

\end{document}

### 2022-11-15 Manuscript/paper.tex
added data section no visuals

@@
 
-Use this section to describe the data that helps to answer your research
-questions.
+The data is acquired from kaggle and is the “largest heart disease data set available”.
+The creator of the data set combined a total of 5 datasets which had 11 overlapping variables.
+There are 918 observations.
+The variables in the data set are as follows: Age,measured in years, sex, male or female, ChestPainType, TA for typical angina, ATA for atypical angina, NAP for non-anginal pain, and ASY for asymptomatic, RestingBP, resting blood pressure measured in mm Hg, Cholesterol, serum cholesterol measured in mg/dL, FastingBS, fasting blood sugar, a binary variable where 1 equals FastingBS greater than 120 mg/dl and 0 otherwise, RestingECG, a binary variable for resting electrocardiogram results where Normal is Normal and ST represents having an ST wave abnormality, MaxHR, maximum heart rate achieved, Exercise angina, a binary yes or no variable, Oldpeak, a numeric value of ST wave depression, ST_Slope, a categorical variable that places the slope of the peak exercise ST segment of a heartbeat wave into Up for upsloping, flat for flat, and down for downsloping, and lastly HeartDisease, the variable of interest, a binary variable where 1 represents an individual with heart disease and 0 for an individual that does not have heart disease.
+As expected with any dataset some data cleaning and variable visualization was required for developing and fitting the model.
+From some rudimentary data visualization it was found that within this data set more females did not have heart disease than females who did.
+The opposite is evidently true for males, more males had heart disease in the data set than males who did not have heart disease.
+In the Age and Heart Disease comparative boxplot it is shown that the median age for individuals who had heart disease is greater than the individuals who did not.
+In the Cholesterol and Heart Disease comparative boxplots the medians for individuals who did and did not have heart disease seem to be very similar.
+Within the cholesterol data column there are over 100 observations with a value of 0, so when constructing the comparative boxplot the 0 values were removed so as not to extraordinarily skew the results.
+The resting blood pressure and heart disease comparative boxplots keep the same theme.
+The median resting blood pressure for individuals with heart disease is slightly higher than for those without heart disease.
+The fasting blood sugar and heart disease proportional bar chart shows that a very high proportion of the individuals that do not have heart disease also have a fasting blood sugar value of less than 120 mg/dl.
+The bar chart also shows that a large proportion of individuals who have heart disease also have a fasting blood sugar of less than 120 mg/dl.
+Finally this barchart shows that while individuals who have a fasting blood sugar that is greater than 120 mg/dl are in the minority there is a higher proportion of them who have heart disease than not.
+A similar trend seems to follow for resting ECG and heart disease.
+The resting ECG and Heart disease proportional bar chart shows that a majority of individuals both with and without heart disease have normal resting ECG outputs.
+However, the proportion of individuals who have heart disease and an ST wave abnormality is larger than the proportion of individuals who do not have heart disease and still have an ST wave abnormality.
+The max heart rate and heart disease comparative boxplot chart shows that the median max heart rate of individuals without heart disease is higher than that for individuals with heart disease.
+The exercise angina and heart disease comparative proportional bar chart shows that a high proportion of individuals who do not have heart disease also do not have exercise induced angina.
+Additionally, a high proportion of individuals who do have heart disease also have exercise induced angina.
+Finally the comparative proportional barchart of ST Slope and heart disease shows that a high proportion of individuals who do not have heart disease have an upward slope on the ST segment of their peak exercise heart beat wave.
+Conversely a high proportion of individuals who have heart disease have a flat slope on the ST segment of their peak exercise heart beat wave.
+Additionally while in both the group with and without heart disease a downward sloping ST segment of the peak exercise heart beat wave was a very small minority of individuals and there was a higher proportion in individuals who had heart disease compared to those that did not.
+PLEASE NOTE: STILL TRYING TO FIGURE OUT HOW TO OUTPUT THE GRAPHS IN ONE VISUAL SO THEY DONt TAKE UP PAGES AND PAGES OF SPACE
 

### 2022-11-15 Manuscript/paper.tex
added the methods section

@@
 
-Use this section to present the methodologies that will generate results by
-analyzing the data.
+The aim for this paper is to create a classical decision tree machine learning model that given variable inputs can predict whether an individual has heart disease or not.
+Three classical decision trees were created.
+One included all of the variables, one included all variables however the data had been modified to void any individuals that had a cholesterol level of 0 mg/dL.
+The final model created had been modified to include all variables except for chest pain type and ST slope.
+Before any classical decision tree was made the data was split into 2 equal sized sets.
+One served as the training set and the other served as the validation or test set.
+A seed number of 4321 was used for the sample function in r so that results could be reproducible.
+After the classical decision tree was made a determination if the tree needed to be trimmed by finding the smallest tree that had a cross validated error within one standard error of the minimum cross validated error.
+To assess the model fit each model was validated with the validation set created at the beginning and the error rate was calculated by the sum of false negative and false positives over the total number of individuals in the set.
 

### 2022-11-15 Manuscript/paper.tex
added results section no pictures

@@
 \label{sec:resu}
+As stated in the methods section, error rate was used to determine model fit and success.
+The first model created, the one with all variables and all observations included produced an  error rate of .496732.
+This means that 49.67\% of the validation observations were misclassified by the model.
+After producing this model and investigating the cross validation table it was determined that no pruning was needed.
+Since the error rate was very large that motivated data modification and reassessment.
+The next model that was made trimmed the number of observations by if an individual had a Cholesterol level of 0.
+This created a decision tree that needed to be pruned.
+Both the tree with 3 splits and 1 split met the requirements of falling within 1 standard error of the smallest cross validated error.
+The model was trimmed to have 3 splits because 1 split felt a little too simplistic.
+After pruning the tree the error rate was found to be .5201.
+This means that 52.01\% of the validation data was misclassified.
+This ended up being the worst performing model.
+The final model created was modified to not include chest pain type because it is a variable that has 4 different classifications and usually variables that are numeric or can have binary classification work much better.
+Additionally ST slope was also removed because another variable in the data set provides a numeric measurement for this and the ST Slope variable, similar to the chest paint type variable, has 3 different classifications instead of 2.
+This tree also showed qualities which would make it a good candidate for pruning however, after attempting pruning in a bevvy of ways my program would seemingly not do any pruning.
+The error rate for this tree ended up being .4924.
+This means 49.24\% of the data in the validation set was misclassified.
+This model was surprisingly the one with the lowest error rate.
 

### 2022-11-15 Manuscript/paper.tex
added discussion section

@@
 
-What are the main contributions again?
-
-What are the limitations of this study?
-
-What are worth pursuing further in the future?
+In conclusion, the error rates of all three models hovered around 50\%.
+This is not a good result.
+Since the first error rate was so large in the first model with all of the variables I was led to try different models to see if they would improve the error rate.
+While each error rate is not identical I would argue that the differences between them are insignificant.
+I wanted to create a model without the cholesterol values of 0 because if the values reported are truly the serum levels of cholesterol it would be impossible if not extremely unhealthy for these to be 0.
+Originally I thought the creator of the data frame might have switched individuals who maybe didn’t have cholesterol levels to report from NA to 0 just so the usability and number of NA values was 0.
+Additionally after getting 2 bad results I decided to reasses and reflect, go through my notes and think about what could possibly make this model better.
+I found that these models work best with variables that are either numeric where a cut off point can be determined or strictly binary variables.
+This led me to take out the chest pain type because it had 4 buckets instead of 2.
+I also took out the ST slop variable because it seemed redundant and it also had 3 buckets of classification instead of 2.
+This model technically ended up having the smallest error rate but it still was around 50\%.
+Limitations in this paper mainly stem from the data source itself.
+Since there are over 100 individuals with claimed blood serum cholesterol levels of 0 when this is not medically possible I think that data point may have been misrepresented.
+Additionally even though there were a significant amount of variables provided I think the addition of more variables like weight, bmi, exercise frequency, or a quantitative measure of healthy lifestyle might make better less error prone models.
+I also think the method chosen to create the models was limiting.
+Since all of the models have a 50\% error rate I can conclude that classical decision trees might not be the best machine learning model for this data.
+Therefore widening the available model creation methods or at least testing other model creation methods and comparing error rates might be valuable to pursuing in the future.
+Finally while this data set was large and quite comprehensive with more time, money, and resources more individuals could be added to the data set.
+Finally and most importantly, this study as it stands in its current state proves something very important: computers won\textsc{\char13}t be overtaking a doctor\textsc{\char13}s role anytime soon.
+The grueling years of training and studying still prevails over a tiny microchip in a motherboard.
+The results of this study however should not discourage the motivation of trying to come up with statistical models for medical outcomes.
+This paper should merely mark a moment in time.
+In the future as statistical knowledge increases and more patient data is released someday a computer may help a doctor in diagnosis.
 

### 2022-11-15 Manuscript/paper.tex
added title and some notes for grader/prof in abstract

@@
 
-\title{Stats Paper}
+\title{Attempting to predict Heart Disease using a classical decision tree machine learning method }
 \author{Olivia Frillici\\
@@
     Here is the abstract. 
+    SOME NOTES: I do not have my abstract or introduction written yet.
+Also I am aware of the lack of references and pictures, for the first draft really wanted to focus on the writing and attempting to get all of the coding done.
     \end{abstract}

### 2022-11-29 Manuscript/paper.tex
took out place holder line
made it so I could output rough draft

@@
-this will be my paper
 \documentclass[12pt]{article}
@@
     % roadmap
-The rest of the paper is organized as follows.
-The data will be presented in Section~\ref{sec:data}.
-The methods are described in Section~\ref{sec:meth}.
-The results are reported in Section~\ref{sec:resu}.
-A discussion concludes in Section~\ref{sec:disc}.
+%The rest of the paper is organized as follows.
+%The data will be presented in Section~\ref{sec:data}.
+%The methods are described in Section~\ref{sec:meth}.
+%The results are reported in Section~\ref{sec:resu}.
+%A discussion concludes in Section~\ref{sec:disc}.
 
@@
 There are 918 observations.
-The variables in the data set are as follows: Age,measured in years, sex, male or female, ChestPainType, TA for typical angina, ATA for atypical angina, NAP for non-anginal pain, and ASY for asymptomatic, RestingBP, resting blood pressure measured in mm Hg, Cholesterol, serum cholesterol measured in mg/dL, FastingBS, fasting blood sugar, a binary variable where 1 equals FastingBS greater than 120 mg/dl and 0 otherwise, RestingECG, a binary variable for resting electrocardiogram results where Normal is Normal and ST represents having an ST wave abnormality, MaxHR, maximum heart rate achieved, Exercise angina, a binary yes or no variable, Oldpeak, a numeric value of ST wave depression, ST_Slope, a categorical variable that places the slope of the peak exercise ST segment of a heartbeat wave into Up for upsloping, flat for flat, and down for downsloping, and lastly HeartDisease, the variable of interest, a binary variable where 1 represents an individual with heart disease and 0 for an individual that does not have heart disease.
+The variables in the data set are as follows: Age,measured in years, sex, male or female, ChestPainType, TA for typical angina, ATA for atypical angina, NAP for non-anginal pain, and ASY for asymptomatic, RestingBP, resting blood pressure measured in mm Hg, Cholesterol, serum cholesterol measured in mg/dL, FastingBS, fasting blood sugar, a binary variable where 1 equals FastingBS greater than 120 mg/dl and 0 otherwise, RestingECG, a binary variable for resting electrocardiogram results where Normal is Normal and ST represents having an ST wave abnormality, MaxHR, maximum heart rate achieved, Exercise angina, a binary yes or no variable, Oldpeak, a numeric value of ST wave depression, ST\_Slope, a categorical variable that places the slope of the peak exercise ST segment of a heartbeat wave into Up for upsloping, flat for flat, and down for downsloping, and lastly HeartDisease, the variable of interest, a binary variable where 1 represents an individual with heart disease and 0 for an individual that does not have heart disease.
 As expected with any dataset some data cleaning and variable visualization was required for developing and fitting the model.
@@
 Additionally while in both the group with and without heart disease a downward sloping ST segment of the peak exercise heart beat wave was a very small minority of individuals and there was a higher proportion in individuals who had heart disease compared to those that did not.
-PLEASE NOTE: STILL TRYING TO FIGURE OUT HOW TO OUTPUT THE GRAPHS IN ONE VISUAL SO THEY DONt TAKE UP PAGES AND PAGES OF SPACE
+PLEASE NOTE: STILL TRYING TO FIGURE OUT HOW TO OUTPUT THE GRAPHS IN ONE VISUAL SO THEY DONt TAKE UP PAGES AND PAGES OF SPACE.
 
@@
 
-\bibliography{refs}
-\bibliographystyle{mcap}
+%\bibliography{refs}
+%\bibliographystyle{ama}
 

### 2022-12-07 Manuscript/paper.tex
added most variable visualization and edited data section

@@
 As expected with any dataset some data cleaning and variable visualization was required for developing and fitting the model.
-From some rudimentary data visualization it was found that within this data set more females did not have heart disease than females who did.
+\par
+\begin{figure}[tbp]
+  \centering
+  \includegraphics[width=\textwidth]{heartdisease_and_gender.png}
+  \caption{A comparative bar chart showing the distribution of gender and heart disease.}
+  \label{fig:heartdisease_and_gender}
+\end{figure}
+Figure~\ref{fig:heartdisease_and_gender} shows that within this data set more females did not have heart disease than females who did.
 The opposite is evidently true for males, more males had heart disease in the data set than males who did not have heart disease.
-In the Age and Heart Disease comparative boxplot it is shown that the median age for individuals who had heart disease is greater than the individuals who did not.
-In the Cholesterol and Heart Disease comparative boxplots the medians for individuals who did and did not have heart disease seem to be very similar.
+\begin{figure}[tbp]
+  \centering
+  \includegraphics[width=\textwidth]{age_heartdisease.png}
+  \caption{A comparative boxplot showing median age of individuals in the data set with and without heart disease.}
+  \label{fig:age_heartdisease}
+\end{figure}
+In figure~\ref{fig:age_heartdisease} the Age and Heart Disease comparative boxplot shows that the median age for individuals who had heart disease is greater than the individuals who did not. 
+\begin{figure}[tbp]
+  \centering
+  \includegraphics[width=\textwidth]{cholesterol_heartdisease.png}
+  \caption{A comparative boxplot showing median age of individuals in the data set with and without heart disease.}
+  \label{fig:cholesterol_heartdisease}
+\end{figure}
+%CHANGE THE IMAGE PUT IN THE WRONG ONE
+In figure~\ref{fig:cholesterol_heartdisease} the Cholesterol and Heart Disease comparative boxplots the medians for individuals who did and did not have heart disease seem to be very similar.
 Within the cholesterol data column there are over 100 observations with a value of 0, so when constructing the comparative boxplot the 0 values were removed so as not to extraordinarily skew the results.
-The resting blood pressure and heart disease comparative boxplots keep the same theme.
+\begin{figure}[tbp]
+  \centering
+  \includegraphics[width=\textwidth]{RestingBloodPressure_Heartdisease.png}
+  \caption{A comparative boxplot showing resting blood pressure in mmHg of individuals in the data set with and without heart disease.}
+  \label{fig:rbp_heartdisease}
+\end{figure}
+In figure~\ref{rbp_heartdisease} the resting blood pressure and heart disease comparative boxplots keep the same theme as figure~\ref{fig:cholesterol_heartdisease}.
 The median resting blood pressure for individuals with heart disease is slightly higher than for those without heart disease.
-The fasting blood sugar and heart disease proportional bar chart shows that a very high proportion of the individuals that do not have heart disease also have a fasting blood sugar value of less than 120 mg/dl.
+However the difference is so mild it appears insignificant.
+\begin{figure}[tbp]
+  \centering
+  \includegraphics[width=\textwidth]{FastingBloodSugar_HeartDisease.png}
+  \caption{A proportional bar chart that shows frequency of individuals who have a fasting blood sugar value of greater than 120 mg/dl or less than within the groups of individuals who have heart disease and who don't.}
+  \label{fig:FBS_heartdisease}
+\end{figure}
+Figure~\ref{fig:FBS_heartdisease} the fasting blood sugar and heart disease proportional bar chart shows that a very high proportion of the individuals that do not have heart disease also have a fasting blood sugar value of less than 120 mg/dl.
 The bar chart also shows that a large proportion of individuals who have heart disease also have a fasting blood sugar of less than 120 mg/dl.
 Finally this barchart shows that while individuals who have a fasting blood sugar that is greater than 120 mg/dl are in the minority there is a higher proportion of them who have heart disease than not.
-A similar trend seems to follow for resting ECG and heart disease.
-The resting ECG and Heart disease proportional bar chart shows that a majority of individuals both with and without heart disease have normal resting ECG outputs.
-However, the proportion of individuals who have heart disease and an ST wave abnormality is larger than the proportion of individuals who do not have heart disease and still have an ST wave abnormality.
-The max heart rate and heart disease comparative boxplot chart shows that the median max heart rate of individuals without heart disease is higher than that for individuals with heart disease.
-The exercise angina and heart disease comparative proportional bar chart shows that a high proportion of individuals who do not have heart disease also do not have exercise induced angina.
+\begin{figure}[tbp]
+  \centering
+  \includegraphics[width=\textwidth]{RestingECG_HeartDisease.png}
+  \caption{A porportional bar chart that shows the frequency of individuals who have a resting electrocardiogram test result of normal or an elctrocardiogram result depicting an ST wave abnormality within the groups of individuals who have or do not have heart disease.}
+  \label{fig:ECG_heartdisease}
+\end{figure}
+Figure~\ref{fig:ECG_heartdisease} shows that a majority of individuals both with and without heart disease have normal resting ECG outputs.
+This graph shows a similar trend to figure~\ref{fig:FBS_heartdisease} in that while having an ST wave abnormality is clearly in the minority a larger proportion of individuals affected by the abnormality also have heart disease.
+\begin{figure}[tbp]
+  \centering
+  \includegraphics[width=\textwidth]{MaxHR_HeartDisease.png}
+  \caption{A comparative boxplot showing maximum heart rate achieved in beats per minute for individuals in the data set with and without heart disease.}
+  \label{fig:maxHR_heartdisease}
+\end{figure}
+Figure~\ref{fig:maxHR_heartdisease}shows that the median max heart rate of individuals without heart disease is higher than that for individuals with heart disease. 
+\begin{figure}[tbp]
+  \centering
+  \includegraphics[width=\textwidth]{ExerciseInducedAngina_HeartDisease.png}
+  \caption{A proportional bar chart that shows the frequency of individuals who have exercise induced angina within the groups of individuals who have or don't have heart disease.}
+  \label{fig:EIA_heartdisease}
+\end{figure}
+Figure\ref{fig:EIA_heartdisease}shows that a high proportion of individuals who do not have heart disease also do not have exercise induced angina.
 Additionally, a high proportion of individuals who do have heart disease also have exercise induced angina.
-Finally the comparative proportional barchart of ST Slope and heart disease shows that a high proportion of individuals who do not have heart disease have an upward slope on the ST segment of their peak exercise heart beat wave.
+\begin{figure}[tbp]
+  \centering
+  \includegraphics[width=\textwidth]{STSlope_HeartDisease.png}
+  \caption{A proportional barchart that shows the frequency of individuals who have an upward sloping, downward sloping, or flat ST segment of their peak exercise heart beat wave within the groups of individuals who have or don't have deart disease.}
+  \label{fig:STslope_heartdisease}
+\end{figure}
+Finally figure~\ref{fig:STslope_heartdisease} shows that a high proportion of individuals who do not have heart disease have an upward slope on the ST segment of their peak exercise heart beat wave.
 Conversely a high proportion of individuals who have heart disease have a flat slope on the ST segment of their peak exercise heart beat wave.
-Additionally while in both the group with and without heart disease a downward sloping ST segment of the peak exercise heart beat wave was a very small minority of individuals and there was a higher proportion in individuals who had heart disease compared to those that did not.
+Additionally while in both the group with and without heart disease a downward sloping ST segment of the peak exercise heart beat wave was a very small minority of individuals, there was a higher proportion in individuals who had heart disease compared to those that did not.
 PLEASE NOTE: STILL TRYING TO FIGURE OUT HOW TO OUTPUT THE GRAPHS IN ONE VISUAL SO THEY DONt TAKE UP PAGES AND PAGES OF SPACE.

### 2022-12-08 Manuscript/paper.tex
added intro and abstract

@@
 \begin{abstract}
-    Here is the abstract. 
-    SOME NOTES: I do not have my abstract or introduction written yet.
-Also I am aware of the lack of references and pictures, for the first draft really wanted to focus on the writing and attempting to get all of the coding done.
+  Heart disease has been the leading cause of death globally for 20 years \citep{}.
+The extraordinary availability of patient data in combination with the potential for great impact derived from successful statistical models serves as motivation for data analysis and model creation.
+The goal of this paper is to create a classical decision tree model using the largest heart disease data set available to predict heart disease.
+A classical decision tree with all variables in the data set had an error rate of .4967, a second model where individuals with a serum cholesterol level of 0 mg/dl were removed had an error rate of .5201, and a third model where the variable chest pain type was removed provided an error rate of .4924.
+The results provided by this paper reconfirm that classical decision trees are not the best model making method for heart disease.
     \end{abstract}
@@
 
-    Use this section to answer three questions:
-    Why is the topic important/interesting?
-    What has been done on this topic in the literature?
-    What is your contribution?
+    To continue living a human’s heart needs to pump blood.
+Blood carries oxygen to our vital organs and also can take away the trash that cells produce as they do their daily tasks.
+Heart disease makes it hard for the human body to function properly.
+By researching and developing tools in which medical providers can input patient data and get outputs pertaining to if a patient has heart disease, medical providers can prescribe lifestyle changes to mitigate and decrease risk of significant cardiac events.
+1.72\% of the world’s population, which is about 126 million individuals, are affected by heart disease \citep{}.
+Additionally 9 million deaths were caused by heart disease globally which makes it the leading cause of death and it has retained this status for the last 20 years\citep{}.
+Heart disease is not only extraordinarily prevalent it is also a quite expensive disease to deal with.
+In 2010 it cost 863 billion us dollars on a global scale \citep{}.
+The prevalence of heart disease, in conjunction with the increased availability of anonymous patient data has led to many people attempting to create models which based on a set of input variables can predict heart disease.
+An example of such work is \citet{} where they compared performances of 3 different models including a decision tree, their tree model ended up having a .21 error rate.
+In this paper a multitude of classical decision trees were grown with the largest heart disease data set available to explore the error rate and if certain data and model manipulation could improve the error rate.
 

### 2022-12-08 Manuscript/paper.tex
added actual citation names and did some light editing

@@
 \begin{abstract}
-  Heart disease has been the leading cause of death globally for 20 years \citep{}.
+  Heart disease has been the leading cause of death globally for 20 years \citep{khan2020global}.
 The extraordinary availability of patient data in combination with the potential for great impact derived from successful statistical models serves as motivation for data analysis and model creation.
@@
 By researching and developing tools in which medical providers can input patient data and get outputs pertaining to if a patient has heart disease, medical providers can prescribe lifestyle changes to mitigate and decrease risk of significant cardiac events.
-1.72\% of the world’s population, which is about 126 million individuals, are affected by heart disease \citep{}.
-Additionally 9 million deaths were caused by heart disease globally which makes it the leading cause of death and it has retained this status for the last 20 years\citep{}.
+1.72\% of the world’s population, which is about 126 million individuals, are affected by heart disease \citep{khan2020global}.
+Additionally 9 million deaths were caused by heart disease globally which makes it the leading cause of death and it has retained this status for the last 20 years\citep{khan2020global}.
 Heart disease is not only extraordinarily prevalent it is also a quite expensive disease to deal with.
-In 2010 it cost 863 billion us dollars on a global scale \citep{}.
+In 2010 it cost 863 billion us dollars on a global scale \citep{khan2020global}.
 The prevalence of heart disease, in conjunction with the increased availability of anonymous patient data has led to many people attempting to create models which based on a set of input variables can predict heart disease.
-An example of such work is \citet{} where they compared performances of 3 different models including a decision tree, their tree model ended up having a .21 error rate.
+An example of such work is \citet{5643666} where they compared performances of 3 different models including a decision tree, their tree model ended up having a .21 error rate.
 In this paper a multitude of classical decision trees were grown with the largest heart disease data set available to explore the error rate and if certain data and model manipulation could improve the error rate.
@@
 Additionally while in both the group with and without heart disease a downward sloping ST segment of the peak exercise heart beat wave was a very small minority of individuals, there was a higher proportion in individuals who had heart disease compared to those that did not.
-PLEASE NOTE: STILL TRYING TO FIGURE OUT HOW TO OUTPUT THE GRAPHS IN ONE VISUAL SO THEY DONt TAKE UP PAGES AND PAGES OF SPACE.
 
@@
 Three classical decision trees were created.
-One included all of the variables, one included all variables however the data had been modified to void any individuals that had a cholesterol level of 0 mg/dL.
-The final model created had been modified to include all variables except for chest pain type and ST slope.
-Before any classical decision tree was made the data was split into 2 equal sized sets.
-One served as the training set and the other served as the validation or test set.
+One included all of the variables and all individuals, one included all variables however the data has been modified to void any individuals that had a cholesterol level of 0 mg/dL.
+The final model created has been modified to include all variables except for chest pain type and ST slope.
+Before any classical decision trees were made the data was split into 2 equal sized sets.
+One served as the training set and the other served as the validation set.
 A seed number of 4321 was used for the sample function in r so that results could be reproducible.
-After the classical decision tree was made a determination if the tree needed to be trimmed by finding the smallest tree that had a cross validated error within one standard error of the minimum cross validated error.
+After the classical decision tree the decision regarding trimming the tree was made a by finding the smallest tree that had a cross validated error within one standard error of the minimum cross validated error.
 To assess the model fit each model was validated with the validation set created at the beginning and the error rate was calculated by the sum of false negative and false positives over the total number of individuals in the set.
@@
 This ended up being the worst performing model.
-The final model created was modified to not include chest pain type because it is a variable that has 4 different classifications and usually variables that are numeric or can have binary classification work much better.
-Additionally ST slope was also removed because another variable in the data set provides a numeric measurement for this and the ST Slope variable, similar to the chest paint type variable, has 3 different classifications instead of 2.
+The final model created was modified to not include chest pain type because it is a variable that has 4 different classifications and usually variables that are numeric or have binary classification work much better.
+Additionally ST slope was also removed because another variable, Oldpeak, in the data set provides a numeric measurement for this.
+Also the ST Slope variable, similar to the chest pain type variable, has 3 different classifications instead of 2.
 This tree also showed qualities which would make it a good candidate for pruning however, after attempting pruning in a bevvy of ways my program would seemingly not do any pruning.

### 2022-12-08 Manuscript/paper.tex
fixed data visualization images

@@
   \centering
-  \includegraphics[width=\textwidth]{heartdisease_and_gender.png}
+  \includegraphics[width=\textwidth]{Sex_HeartDisease.png}
   \caption{A comparative bar chart showing the distribution of gender and heart disease.}
-  \label{fig:heartdisease_and_gender}
+  \label{fig:heartdisease_and_sex}
 \end{figure}
-Figure~\ref{fig:heartdisease_and_gender} shows that within this data set more females did not have heart disease than females who did.
+Figure~\ref{fig:heartdisease_and_sex} shows that within this data set more females did not have heart disease than females who did.
 The opposite is evidently true for males, more males had heart disease in the data set than males who did not have heart disease.
@@
   \centering
-  \includegraphics[width=\textwidth]{age_heartdisease.png}
+  \includegraphics[width=\textwidth]{Age_HeartDisease.png}
   \caption{A comparative boxplot showing median age of individuals in the data set with and without heart disease.}
@@
   \centering
-  \includegraphics[width=\textwidth]{cholesterol_heartdisease.png}
-  \caption{A comparative boxplot showing median age of individuals in the data set with and without heart disease.}
+  \includegraphics[width=\textwidth]{Cholesterol_Heart Disease.png}
+  \caption{A comparative boxplot showing serum cholesterol level mg/dL for individuals in the data set with and without heart disease.}
   \label{fig:cholesterol_heartdisease}
 \end{figure}
-%CHANGE THE IMAGE PUT IN THE WRONG ONE
 In figure~\ref{fig:cholesterol_heartdisease} the Cholesterol and Heart Disease comparative boxplots the medians for individuals who did and did not have heart disease seem to be very similar.
@@
   \centering
-  \includegraphics[width=\textwidth]{RestingBloodPressure_Heartdisease.png}
+  \includegraphics[width=\textwidth]{RBP_HeartDisease.png}
   \caption{A comparative boxplot showing resting blood pressure in mmHg of individuals in the data set with and without heart disease.}
@@
 
-%\bibliography{refs}
-%\bibliographystyle{ama}
+\bibliography{refs}
+\bibliographystyle{ama}
 

### 2022-12-09 Manuscript/paper.tex
added rest of pictures edited discussion and conclusion

@@
 \end{figure}
-Figure~\ref{fig:heartdisease_and_sex} shows that within this data set more females did not have heart disease than females who did.
+Figure~\ref*{fig:heartdisease_and_sex} shows that within this data set more females did not have heart disease than females who did.
 The opposite is evidently true for males, more males had heart disease in the data set than males who did not have heart disease.
@@
 \end{figure}
-In figure~\ref{fig:age_heartdisease} the Age and Heart Disease comparative boxplot shows that the median age for individuals who had heart disease is greater than the individuals who did not. 
+In figure~\ref*{fig:age_heartdisease} the Age and Heart Disease comparative boxplot shows that the median age for individuals who had heart disease is greater than the individuals who did not. 
 \begin{figure}[tbp]
@@
 \end{figure}
-In figure~\ref{fig:cholesterol_heartdisease} the Cholesterol and Heart Disease comparative boxplots the medians for individuals who did and did not have heart disease seem to be very similar.
+In figure~\ref*{fig:cholesterol_heartdisease} the Cholesterol and Heart Disease comparative boxplots the medians for individuals who did and did not have heart disease seem to be very similar.
 Within the cholesterol data column there are over 100 observations with a value of 0, so when constructing the comparative boxplot the 0 values were removed so as not to extraordinarily skew the results.
@@
 \end{figure}
-In figure~\ref{rbp_heartdisease} the resting blood pressure and heart disease comparative boxplots keep the same theme as figure~\ref{fig:cholesterol_heartdisease}.
+In figure~\ref*{rbp_heartdisease} the resting blood pressure and heart disease comparative boxplots keep the same theme as figure~\ref*{fig:cholesterol_heartdisease}.
 The median resting blood pressure for individuals with heart disease is slightly higher than for those without heart disease.
@@
 \end{figure}
-Figure~\ref{fig:FBS_heartdisease} the fasting blood sugar and heart disease proportional bar chart shows that a very high proportion of the individuals that do not have heart disease also have a fasting blood sugar value of less than 120 mg/dl.
+Figure~\ref*{fig:FBS_heartdisease} the fasting blood sugar and heart disease proportional bar chart shows that a very high proportion of the individuals that do not have heart disease also have a fasting blood sugar value of less than 120 mg/dl.
 The bar chart also shows that a large proportion of individuals who have heart disease also have a fasting blood sugar of less than 120 mg/dl.
@@
 \end{figure}
-Figure~\ref{fig:ECG_heartdisease} shows that a majority of individuals both with and without heart disease have normal resting ECG outputs.
-This graph shows a similar trend to figure~\ref{fig:FBS_heartdisease} in that while having an ST wave abnormality is clearly in the minority a larger proportion of individuals affected by the abnormality also have heart disease.
+Figure~\ref*{fig:ECG_heartdisease} shows that a majority of individuals both with and without heart disease have normal resting ECG outputs.
+This graph shows a similar trend to figure~\ref*{fig:FBS_heartdisease} in that while having an ST wave abnormality is clearly in the minority a larger proportion of individuals affected by the abnormality also have heart disease.
 \begin{figure}[tbp]
@@
 \end{figure}
-Figure~\ref{fig:maxHR_heartdisease}shows that the median max heart rate of individuals without heart disease is higher than that for individuals with heart disease. 
+Figure~\ref*{fig:maxHR_heartdisease}shows that the median max heart rate of individuals without heart disease is higher than that for individuals with heart disease. 
 \begin{figure}[tbp]
@@
 \end{figure}
-Figure\ref{fig:EIA_heartdisease}shows that a high proportion of individuals who do not have heart disease also do not have exercise induced angina.
+Figure\ref*{fig:EIA_heartdisease}shows that a high proportion of individuals who do not have heart disease also do not have exercise induced angina.
 Additionally, a high proportion of individuals who do have heart disease also have exercise induced angina.
@@
 \end{figure}
-Finally figure~\ref{fig:STslope_heartdisease} shows that a high proportion of individuals who do not have heart disease have an upward slope on the ST segment of their peak exercise heart beat wave.
+Finally figure~\ref*{fig:STslope_heartdisease} shows that a high proportion of individuals who do not have heart disease have an upward slope on the ST segment of their peak exercise heart beat wave.
 Conversely a high proportion of individuals who have heart disease have a flat slope on the ST segment of their peak exercise heart beat wave.
 Additionally while in both the group with and without heart disease a downward sloping ST segment of the peak exercise heart beat wave was a very small minority of individuals, there was a higher proportion in individuals who had heart disease compared to those that did not.
+\par
 
@@
 One included all of the variables and all individuals, one included all variables however the data has been modified to void any individuals that had a cholesterol level of 0 mg/dL.
-The final model created has been modified to include all variables except for chest pain type and ST slope.
+The final model created had been modified to include all variables except for chest pain type and ST slope.
+\par
 Before any classical decision trees were made the data was split into 2 equal sized sets.
@@
 A seed number of 4321 was used for the sample function in r so that results could be reproducible.
-After the classical decision tree the decision regarding trimming the tree was made a by finding the smallest tree that had a cross validated error within one standard error of the minimum cross validated error.
+\par
+After the classical decision tree the decision regarding trimming the tree was made a by plotting the cross validated error against the complexity parameter.
+Trees considered for the final model had points below the horizontal line on the cp graph.
+\par
 To assess the model fit each model was validated with the validation set created at the beginning and the error rate was calculated by the sum of false negative and false positives over the total number of individuals in the set.
+\par
 
@@
 This means that 49.67\% of the validation observations were misclassified by the model.
-After producing this model and investigating the cross validation table it was determined that no pruning was needed.
-Since the error rate was very large that motivated data modification and reassessment.
-The next model that was made trimmed the number of observations by if an individual had a Cholesterol level of 0.
-This created a decision tree that needed to be pruned.
-Both the tree with 3 splits and 1 split met the requirements of falling within 1 standard error of the smallest cross validated error.
+\begin{figure}[tbp]
+  \centering
+  \includegraphics[width=\textwidth]{cpGraph_AllVars.png}
+  \caption{Cross validationerror versus the complexity parameter graph for the model developed with all variables and all individuals included.}
+  \label{fig:cpGraph_AllVars}
+\end{figure}
+After producing this model and investigating figure~\ref*{cpGraph_AllVars},the cross validation error versus the complexity parameter graph, it was determined that no pruning was needed.
+No pruning was needed because as shown in figure~\ref*{cpGraph_AllVars} all of the different trees fall below the horizontal line plotted on the graph.
+Large error rate motivated data modification and reassessment.
+\par
+The next model that was made trimmed the number of observations by if an individual had a serum cholesterol level of 0 mg/dL. 
+\begin{figure}[tbp]
+  \centering
+  \includegraphics[width=\textwidth]{cpGraph_NoCholesterol.png}
+  \caption{Cross validation error versus the complexity parameter graph for the model developed with no individuals that had a serum cholesterol level of 0 mg/dL.}
+  \label{fig:cpGraph_NoCholesterol}
+\end{figure}
+As shown in figure~\ref*{fig:cpGraph_NoCholesterol} it is seen that a tree with 2 terminal nodes,4 terminal nodes, 6 terminal nodes, 7 terminal nodes, and 11 terminal nodes all fit within the cirterion of having a relative error that is below the horizontal line plotted on the graph.
+Generally a smaller tree is reckognized as better because the computer has to do less work.
+Therefore the two sincere contenders for this model are the tree with 2 terminal nodes and one split or the tree with 4 terminal nodes and 3 splits.
 The model was trimmed to have 3 splits because 1 split felt a little too simplistic.
@@
 This means that 52.01\% of the validation data was misclassified.
-This ended up being the worst performing model.
+This ended up being the worst performing model with the greatest error rate.
+\par
 The final model created was modified to not include chest pain type because it is a variable that has 4 different classifications and usually variables that are numeric or have binary classification work much better.
@@
 Also the ST Slope variable, similar to the chest pain type variable, has 3 different classifications instead of 2.
-This tree also showed qualities which would make it a good candidate for pruning however, after attempting pruning in a bevvy of ways my program would seemingly not do any pruning.
-The error rate for this tree ended up being .4924.
+\begin{figure}[tbp]
+  \centering
+  \includegraphics[width=\textwidth]{cpGraph_NoSTSandCPT.png}
+  \caption{Cross validation error versus the complexity parameter graph for the model developed with no the ST Slope and Chest Pain Type variables removed.}
+  \label{fig:cpGraph_NoSTSandCPT}
+\end{figure} 
+This tree also showed qualities which would make it a good candidate for pruning.
+As seen in figure~\ref*{fig:cpGraph_NoSTSandCPT} it is shown that both a tree with 3 terminal nodes and 2 splits or 6 terminal nodes and 5 splits would be optimal.
+However, after attempting pruning in a bevvy of ways my program would seemingly not do any pruning.
+The error rate for this tree, with 6 terminal nodes and 5 splits, ended up being .4924.
 This means 49.24\% of the data in the validation set was misclassified.
 This model was surprisingly the one with the lowest error rate.
-
+\par
+\begin{figure}[tbp]
+  \centering
+  \includegraphics[width=\textwidth]{DTree_noSTSandCPT.png}
+  \caption{Best performing classical decision tree, with an error rate of .4924, out of the three. This tree was constructed without ST Slope and Chest Pain Type variables. }
+  \label{fig:DTree_noSTSandCPT}
+\end{figure} 
+The final and best decision tree out of the three models constructed in this paer is shown in figure~\ref*{fig:DTree_noSTSandCPT}.
+Exercise Angina, Cholesterol, and Old Peak variables are deemed the most significant at determining heart disease in individuals in this model.
+\par
 \section{Discussion}
@@
 While each error rate is not identical I would argue that the differences between them are insignificant.
-I wanted to create a model without the cholesterol values of 0 because if the values reported are truly the serum levels of cholesterol it would be impossible if not extremely unhealthy for these to be 0.
+I wanted to create a model without the cholesterol values of 0 mg/dL because if the values reported are truly the serum levels of cholesterol it would be impossible if not extremely unhealthy for these to be 0 mg/dL.
 Originally I thought the creator of the data frame might have switched individuals who maybe didn’t have cholesterol levels to report from NA to 0 just so the usability and number of NA values was 0.
@@
 This led me to take out the chest pain type because it had 4 buckets instead of 2.
-I also took out the ST slop variable because it seemed redundant and it also had 3 buckets of classification instead of 2.
+I also took out the ST slope variable because it seemed redundant and it also had 3 buckets of classification instead of 2.
 This model technically ended up having the smallest error rate but it still was around 50\%.
+\par
 Limitations in this paper mainly stem from the data source itself.
-Since there are over 100 individuals with claimed blood serum cholesterol levels of 0 when this is not medically possible I think that data point may have been misrepresented.
-Additionally even though there were a significant amount of variables provided I think the addition of more variables like weight, bmi, exercise frequency, or a quantitative measure of healthy lifestyle might make better less error prone models.
+Since there are over 100 individuals with claimed blood serum cholesterol levels of 0 mg/dL when this is not medically possible I think that data point may have been misrepresented.
+Additionally even though there were a significant amount of variables provided I think the addition of more variables like weight, BMI, exercise frequency, or a quantitative measure of healthy lifestyle might make better less error prone models.
 I also think the method chosen to create the models was limiting.
-Since all of the models have a 50\% error rate I can conclude that classical decision trees might not be the best machine learning model for this data.
-Therefore widening the available model creation methods or at least testing other model creation methods and comparing error rates might be valuable to pursuing in the future.
-Finally while this data set was large and quite comprehensive with more time, money, and resources more individuals could be added to the data set.
+Since all of the models have a 50\% error rate I can conclude and reconfirm that classical decision trees are not be the best machine learning model for this data.
+At the time of writing the proposal for this project and then when developing the rough draft and doing the actual coding and model creation for this project I did not have the knowledge of many other methods of machine learning for binary variables at the time.
+As the semester as continued and my knowledge from other classes has expanded the breadth of statisitical idease that I know, it would be incredibly beneficial to delve into models such as logistic regression, linear discriminant analysis, quadratic discriminant analysis, and deep learning models such as feed foward networks.
+Finally while this data set was large and quite comprehensive with more time, money, and resources more individuals and variables could be added to the data set.
+\par
 Finally and most importantly, this study as it stands in its current state proves something very important: computers won\textsc{\char13}t be overtaking a doctor\textsc{\char13}s role anytime soon.
@@
 This paper should merely mark a moment in time.
-In the future as statistical knowledge increases and more patient data is released someday a computer may help a doctor in diagnosis.
+In the future as statistical knowledge increases, as more patient data is released, and as computers get better, faster, and smarter, someday a computer may help a doctor in diagnosis of heart disease.
 

### 2022-12-09 Manuscript/paper.tex
did some light final editing

@@
 
-% for space filling
-\usepackage{lipsum}
 % highlighting hyper links
@@
 
-\title{Attempting to predict Heart Disease using a classical decision tree machine learning method }
+\title{Attempting to Predict Heart Disease using a Classical Decision Tree Machine Learning Method }
 \author{Olivia Frillici\\
@@
 \begin{abstract}
-  Heart disease has been the leading cause of death globally for 20 years \citep{khan2020global}.
+  Heart disease has been the leading cause of death globally for 20 years \citep*{khan2020global}.
 The extraordinary availability of patient data in combination with the potential for great impact derived from successful statistical models serves as motivation for data analysis and model creation.
-The goal of this paper is to create a classical decision tree model using the largest heart disease data set available to predict heart disease.
-A classical decision tree with all variables in the data set had an error rate of .4967, a second model where individuals with a serum cholesterol level of 0 mg/dl were removed had an error rate of .5201, and a third model where the variable chest pain type was removed provided an error rate of .4924.
+The goal of this paper is to create a classical decision tree model using the largest heart disease data set available to predict heart disease \citep*{CitekeyMisc}.
+A classical decision tree with all variables in the data set had an error rate of .4967, a second model where individuals with a serum cholesterol level of 0 mg/dl were removed had an error rate of .5201, and a third model where the variable chest pain type and ST Slope were removed provided an error rate of .4924.
 The results provided by this paper reconfirm that classical decision trees are not the best model making method for heart disease.
     \end{abstract}
+  \newpage
 
@@
 
-The data is acquired from kaggle and is the “largest heart disease data set available”.
+The data is acquired from kaggle and is the “largest heart disease data set available” \citep*{CitekeyMisc}.
 The creator of the data set combined a total of 5 datasets which had 11 overlapping variables.
@@
   \includegraphics[width=\textwidth]{RestingECG_HeartDisease.png}
-  \caption{A porportional bar chart that shows the frequency of individuals who have a resting electrocardiogram test result of normal or an elctrocardiogram result depicting an ST wave abnormality within the groups of individuals who have or do not have heart disease.}
+  \caption{A proportional bar chart that shows the frequency of individuals who have a resting electrocardiogram test result of normal or an elctrocardiogram result depicting an ST wave abnormality within the groups of individuals who have or do not have heart disease.}
   \label{fig:ECG_heartdisease}
@@
 At the time of writing the proposal for this project and then when developing the rough draft and doing the actual coding and model creation for this project I did not have the knowledge of many other methods of machine learning for binary variables at the time.
-As the semester as continued and my knowledge from other classes has expanded the breadth of statisitical idease that I know, it would be incredibly beneficial to delve into models such as logistic regression, linear discriminant analysis, quadratic discriminant analysis, and deep learning models such as feed foward networks.
+As the semester has continued and my knowledge from other classes has expanded the breadth of statisitical processes that I know, it would be incredibly beneficial to delve into models such as logistic regression, linear discriminant analysis, quadratic discriminant analysis, and deep learning models such as feed foward networks.
 Finally while this data set was large and quite comprehensive with more time, money, and resources more individuals and variables could be added to the data set.

### 2022-12-09 Manuscript/paper.tex
replaced apostrophes with the escape format

@@
 
-    To continue living a human’s heart needs to pump blood.
+    To continue living a human\textsc{\char13}s heart needs to pump blood.
 Blood carries oxygen to our vital organs and also can take away the trash that cells produce as they do their daily tasks.
@@
 By researching and developing tools in which medical providers can input patient data and get outputs pertaining to if a patient has heart disease, medical providers can prescribe lifestyle changes to mitigate and decrease risk of significant cardiac events.
-1.72\% of the world’s population, which is about 126 million individuals, are affected by heart disease \citep{khan2020global}.
+1.72\% of the world\textsc{\char13}s population, which is about 126 million individuals, are affected by heart disease \citep{khan2020global}.
 Additionally 9 million deaths were caused by heart disease globally which makes it the leading cause of death and it has retained this status for the last 20 years\citep{khan2020global}.
@@
 I wanted to create a model without the cholesterol values of 0 mg/dL because if the values reported are truly the serum levels of cholesterol it would be impossible if not extremely unhealthy for these to be 0 mg/dL.
-Originally I thought the creator of the data frame might have switched individuals who maybe didn’t have cholesterol levels to report from NA to 0 just so the usability and number of NA values was 0.
+Originally I thought the creator of the data frame might have switched individuals who maybe didn\textsc{\char13}t have cholesterol levels to report from NA to 0 just so the usability and number of NA values was 0.
 Additionally after getting 2 bad results I decided to reasses and reflect, go through my notes and think about what could possibly make this model better.

### 2022-12-09 Manuscript/paper.tex
fixed a few more errors

@@
 At the time of writing the proposal for this project and then when developing the rough draft and doing the actual coding and model creation for this project I did not have the knowledge of many other methods of machine learning for binary variables at the time.
-As the semester has continued and my knowledge from other classes has expanded the breadth of statisitical processes that I know, it would be incredibly beneficial to delve into models such as logistic regression, linear discriminant analysis, quadratic discriminant analysis, and deep learning models such as feed foward networks.
+As the semester has continued and my knowledge from other classes has expanded the breadth of statistical processes that I know, it would be incredibly beneficial to delve into models such as logistic regression, linear discriminant analysis, quadratic discriminant analysis, and deep learning models such as feed forward networks.
 Finally while this data set was large and quite comprehensive with more time, money, and resources more individuals and variables could be added to the data set.

### 2022-12-09 Manuscript/paper.tex
reconfigured the variable visualization graphs more light editing and typesetting to ensure to hit page length

@@
 \usepackage{natbib}
+\usepackage{caption}
+\usepackage{subcaption}
+\usepackage{setspace}
+\doublespacing
 
@@
     \end{abstract}
+
   \newpage
@@
 By researching and developing tools in which medical providers can input patient data and get outputs pertaining to if a patient has heart disease, medical providers can prescribe lifestyle changes to mitigate and decrease risk of significant cardiac events.
-1.72\% of the world\textsc{\char13}s population, which is about 126 million individuals, are affected by heart disease \citep{khan2020global}.
-Additionally 9 million deaths were caused by heart disease globally which makes it the leading cause of death and it has retained this status for the last 20 years\citep{khan2020global}.
+1.72\% of the world\textsc{\char13}s population, which is about 126 million individuals, are affected by heart disease \citep*{khan2020global}.
+Additionally 9 million deaths were caused by heart disease globally which makes it the leading cause of death and it has retained this status for the last 20 years\citep*{khan2020global}.
 Heart disease is not only extraordinarily prevalent it is also a quite expensive disease to deal with.
-In 2010 it cost 863 billion us dollars on a global scale \citep{khan2020global}.
+In 2010 it cost 863 billion us dollars on a global scale \citep*{khan2020global}.
 The prevalence of heart disease, in conjunction with the increased availability of anonymous patient data has led to many people attempting to create models which based on a set of input variables can predict heart disease.
-An example of such work is \citet{5643666} where they compared performances of 3 different models including a decision tree, their tree model ended up having a .21 error rate.
-In this paper a multitude of classical decision trees were grown with the largest heart disease data set available to explore the error rate and if certain data and model manipulation could improve the error rate.
+An example of such work is \citet*{5643666} where they compared performances of 3 different models including a decision tree, which ended up having a .21 error rate.
+In this paper a multitude of classical decision trees were grown with the largest heart disease data set available to explore the error rate and if certain data and model manipulation could improve the error rate.\par
 
-    % roadmap
-%The rest of the paper is organized as follows.
-%The data will be presented in Section~\ref{sec:data}.
-%The methods are described in Section~\ref{sec:meth}.
-%The results are reported in Section~\ref{sec:resu}.
-%A discussion concludes in Section~\ref{sec:disc}.
+The rest of the paper is organized as follows. 
+Then the data will be presented in Section~\ref{sec:data}.
+The methods are described in Section~\ref{sec:meth}.
+The results are reported in Section~\ref{sec:resu}.
+and discussion concludes in Section~\ref{sec:disc}.
 
@@
 \label{sec:data}
+\begin{figure}
+  \centering
+  \begin{subfigure}[b]{0.3\textwidth}
+       \centering
+       \includegraphics[width=\textwidth]{Sex_HeartDisease.png}
+       \caption{A comparative bar chart showing the distribution of sex and heart disease.}
+       \label{fig:heartdisease_and_sex}
+   \end{subfigure}
+   \hfill
+   \begin{subfigure}[b]{0.3\textwidth}
+       \centering
+       \includegraphics[width=\textwidth]{age_heartdisease.png}
+       \caption{A comparative boxplot showing median age of individuals in the data set with and without heart disease.}
+       \label{fig:age_heartdisease}
+   \end{subfigure}
+   \hfill
+   \begin{subfigure}[b]{0.3\textwidth}
+       \centering
+       \includegraphics[width=\textwidth]{Cholesterol_Heart Disease.png}
+       \caption{A comparative boxplot showing serum cholesterol level mg/dL for individuals in the data set with and without heart disease.}
+       \label{fig:cholesterol_heartdisease}
+   \end{subfigure}
 
+   \begin{subfigure}[b]{0.3\textwidth}
+       \centering
+       \includegraphics[width=\textwidth]{RBP_HeartDisease.png}
+       \caption{A comparative boxplot showing resting blood pressure in mmHg of individuals in the data set with and without heart disease.}
+       \label{fig:rbp_heartdisease}
+   \end{subfigure}
+   \hfill
+   \begin{subfigure}[b]{0.3\textwidth}
+       \centering
+       \includegraphics[width=\textwidth]{FastingBloodSugar_HeartDisease.png}
+       \caption{A proportional bar chart that shows frequency of individuals who have a fasting blood sugar value of greater than 120 mg/dl or less than within the groups of individuals who have heart disease and who don't.}
+      \label{fig:FBS_heartdisease}
+   \end{subfigure}
+   \hfill
+   \begin{subfigure}[b]{0.3\textwidth}
+       \centering
+       \includegraphics[width=\textwidth]{RestingECG_HeartDisease.png}
+       \caption{A proportional bar chart that shows the frequency of individuals who have a resting electrocardiogram test result of normal or an elctrocardiogram result depicting an ST wave abnormality within the groups of individuals who have or do not have heart disease.}
+      \label{fig:ECG_heartdisease}
+   \end{subfigure}
+
+   \begin{subfigure}[b]{0.3\textwidth}
+       \centering
+       \includegraphics[width=\textwidth]{MaxHR_HeartDisease.png}
+       \caption{A comparative boxplot showing maximum heart rate achieved in beats per minute for individuals in the data set with and without heart disease.}
+       \label{fig:maxHR_heartdisease}
+   \end{subfigure}
+   \hfill
+   \begin{subfigure}[b]{0.3\textwidth}
+       \centering
+       \includegraphics[width=\textwidth]{ExerciseInducedAngina_HeartDisease.png}
+       \caption{A proportional bar chart that shows the frequency of individuals who have exercise induced angina within the groups of individuals who have or don't have heart disease.}
+       \label{fig:EIA_heartdisease}
+   \end{subfigure}
+   \hfill
+   \begin{subfigure}[b]{0.3\textwidth}
+       \centering
+       \includegraphics[width=\textwidth]{STSlope_HeartDisease.png}
+       \caption{A proportional barchart that shows the frequency of individuals who have an upward sloping, downward sloping, or flat ST segment of their peak exercise heart beat wave within the groups of individuals who have or don't have deart disease.}
+       \label{fig:STslope_heartdisease}
+   \end{subfigure}
+      \caption{Data visualization of the input variables}
+      \label{fig:datavis}
+\end{figure}
 The data is acquired from kaggle and is the “largest heart disease data set available” \citep*{CitekeyMisc}.
@@
 The variables in the data set are as follows: Age,measured in years, sex, male or female, ChestPainType, TA for typical angina, ATA for atypical angina, NAP for non-anginal pain, and ASY for asymptomatic, RestingBP, resting blood pressure measured in mm Hg, Cholesterol, serum cholesterol measured in mg/dL, FastingBS, fasting blood sugar, a binary variable where 1 equals FastingBS greater than 120 mg/dl and 0 otherwise, RestingECG, a binary variable for resting electrocardiogram results where Normal is Normal and ST represents having an ST wave abnormality, MaxHR, maximum heart rate achieved, Exercise angina, a binary yes or no variable, Oldpeak, a numeric value of ST wave depression, ST\_Slope, a categorical variable that places the slope of the peak exercise ST segment of a heartbeat wave into Up for upsloping, flat for flat, and down for downsloping, and lastly HeartDisease, the variable of interest, a binary variable where 1 represents an individual with heart disease and 0 for an individual that does not have heart disease.
-As expected with any dataset some data cleaning and variable visualization was required for developing and fitting the model.
+As expected with any data set some data cleaning and variable visualization was required for developing and fitting the model.
 \par
-\begin{figure}[tbp]
-  \centering
-  \includegraphics[width=\textwidth]{Sex_HeartDisease.png}
-  \caption{A comparative bar chart showing the distribution of gender and heart disease.}
-  \label{fig:heartdisease_and_sex}
-\end{figure}
+
 Figure~\ref*{fig:heartdisease_and_sex} shows that within this data set more females did not have heart disease than females who did.
 The opposite is evidently true for males, more males had heart disease in the data set than males who did not have heart disease.
-\begin{figure}[tbp]
-  \centering
-  \includegraphics[width=\textwidth]{Age_HeartDisease.png}
-  \caption{A comparative boxplot showing median age of individuals in the data set with and without heart disease.}
-  \label{fig:age_heartdisease}
-\end{figure}
+
 In figure~\ref*{fig:age_heartdisease} the Age and Heart Disease comparative boxplot shows that the median age for individuals who had heart disease is greater than the individuals who did not.
-\begin{figure}[tbp]
-  \centering
-  \includegraphics[width=\textwidth]{Cholesterol_Heart Disease.png}
-  \caption{A comparative boxplot showing serum cholesterol level mg/dL for individuals in the data set with and without heart disease.}
-  \label{fig:cholesterol_heartdisease}
-\end{figure}
+
 In figure~\ref*{fig:cholesterol_heartdisease} the Cholesterol and Heart Disease comparative boxplots the medians for individuals who did and did not have heart disease seem to be very similar.
 Within the cholesterol data column there are over 100 observations with a value of 0, so when constructing the comparative boxplot the 0 values were removed so as not to extraordinarily skew the results.
-\begin{figure}[tbp]
-  \centering
-  \includegraphics[width=\textwidth]{RBP_HeartDisease.png}
-  \caption{A comparative boxplot showing resting blood pressure in mmHg of individuals in the data set with and without heart disease.}
-  \label{fig:rbp_heartdisease}
-\end{figure}
-In figure~\ref*{rbp_heartdisease} the resting blood pressure and heart disease comparative boxplots keep the same theme as figure~\ref*{fig:cholesterol_heartdisease}.
+
+In figure~\ref*{fig:rbp_heartdisease} the resting blood pressure and heart disease comparative boxplots keep the same theme as figure~\ref*{fig:cholesterol_heartdisease}.
 The median resting blood pressure for individuals with heart disease is slightly higher than for those without heart disease.
 However the difference is so mild it appears insignificant.
-\begin{figure}[tbp]
-  \centering
-  \includegraphics[width=\textwidth]{FastingBloodSugar_HeartDisease.png}
-  \caption{A proportional bar chart that shows frequency of individuals who have a fasting blood sugar value of greater than 120 mg/dl or less than within the groups of individuals who have heart disease and who don't.}
-  \label{fig:FBS_heartdisease}
-\end{figure}
+
 Figure~\ref*{fig:FBS_heartdisease} the fasting blood sugar and heart disease proportional bar chart shows that a very high proportion of the individuals that do not have heart disease also have a fasting blood sugar value of less than 120 mg/dl.
@@
 Finally this barchart shows that while individuals who have a fasting blood sugar that is greater than 120 mg/dl are in the minority there is a higher proportion of them who have heart disease than not.
-\begin{figure}[tbp]
-  \centering
-  \includegraphics[width=\textwidth]{RestingECG_HeartDisease.png}
-  \caption{A proportional bar chart that shows the frequency of individuals who have a resting electrocardiogram test result of normal or an elctrocardiogram result depicting an ST wave abnormality within the groups of individuals who have or do not have heart disease.}
-  \label{fig:ECG_heartdisease}
-\end{figure}
+
 Figure~\ref*{fig:ECG_heartdisease} shows that a majority of individuals both with and without heart disease have normal resting ECG outputs.
 This graph shows a similar trend to figure~\ref*{fig:FBS_heartdisease} in that while having an ST wave abnormality is clearly in the minority a larger proportion of individuals affected by the abnormality also have heart disease.
-\begin{figure}[tbp]
-  \centering
-  \includegraphics[width=\textwidth]{MaxHR_HeartDisease.png}
-  \caption{A comparative boxplot showing maximum heart rate achieved in beats per minute for individuals in the data set with and without heart disease.}
-  \label{fig:maxHR_heartdisease}
-\end{figure}
+
 Figure~\ref*{fig:maxHR_heartdisease}shows that the median max heart rate of individuals without heart disease is higher than that for individuals with heart disease.
-\begin{figure}[tbp]
-  \centering
-  \includegraphics[width=\textwidth]{ExerciseInducedAngina_HeartDisease.png}
-  \caption{A proportional bar chart that shows the frequency of individuals who have exercise induced angina within the groups of individuals who have or don't have heart disease.}
-  \label{fig:EIA_heartdisease}
-\end{figure}
+The median max heart rate for individuals with no heart disease is approximately 150 beats per minute.
+In contrast individuals with heart disease have a median max heart rate of 130 beats per minute.
+
 Figure\ref*{fig:EIA_heartdisease}shows that a high proportion of individuals who do not have heart disease also do not have exercise induced angina.
 Additionally, a high proportion of individuals who do have heart disease also have exercise induced angina.
-\begin{figure}[tbp]
-  \centering
-  \includegraphics[width=\textwidth]{STSlope_HeartDisease.png}
-  \caption{A proportional barchart that shows the frequency of individuals who have an upward sloping, downward sloping, or flat ST segment of their peak exercise heart beat wave within the groups of individuals who have or don't have deart disease.}
-  \label{fig:STslope_heartdisease}
-\end{figure}
+
 Finally figure~\ref*{fig:STslope_heartdisease} shows that a high proportion of individuals who do not have heart disease have an upward slope on the ST segment of their peak exercise heart beat wave.
 Conversely a high proportion of individuals who have heart disease have a flat slope on the ST segment of their peak exercise heart beat wave.
-Additionally while in both the group with and without heart disease a downward sloping ST segment of the peak exercise heart beat wave was a very small minority of individuals, there was a higher proportion in individuals who had heart disease compared to those that did not.
+Additionally while in both the group with and without heart disease a downward sloping ST segment of the peak exercise heart beat wave was a very small minority of individuals, there was a higher proportion of individuals who had heart disease  affected compared to those that did not.
 \par
@@
 Before any classical decision trees were made the data was split into 2 equal sized sets.
-One served as the training set and the other served as the validation set.
-A seed number of 4321 was used for the sample function in r so that results could be reproducible.
+In totality 2 training sets and 2 training sets were created this is the result of needing a training set and validation set that did not have individuals with cholesterol values of 0 mg/dl.
+The oher training and validation set had all individuals in it.
+A seed number of 4321 was used for the sample function when creating both sets of training and validation sets in r so that results could be reproducible.
 \par
-After the classical decision tree the decision regarding trimming the tree was made a by plotting the cross validated error against the complexity parameter.
-Trees considered for the final model had points below the horizontal line on the cp graph.
+After the classical decision tree was produced the decision regarding trimming the tree was made by plotting the cross validated error against the complexity parameter which produced a complexity parameter graph.
+Trees considered for the final model had points below the dotted horizontal line on the complexity parameter graph.
 \par
-To assess the model fit each model was validated with the validation set created at the beginning and the error rate was calculated by the sum of false negative and false positives over the total number of individuals in the set.
+To assess the model fit the model that included all variables and all individuals and the model that included all individuals but redacted the ST Slope variable and the Chest Pain Type variable was validated with the validation set that was created using all rows of the data.
+The model created that removed all individuals with a cholesterol level equal to 0 mg/dL was validated with a validation set that was created out of the subsetted data.
+For comparison and assessment of model success the error rate was calculated.
+The error rate was found by summing the false negatives and positives and dividing by the total number of individuals in the set.
 \par
@@
 \label{sec:resu}
-As stated in the methods section, error rate was used to determine model fit and success.
-The first model created, the one with all variables and all observations included produced an  error rate of .496732.
+As stated in the methods section, error rate was used to determine model success.
+The first model created, the one with all variables and all individuals included produced an error rate of .496732.
 This means that 49.67\% of the validation observations were misclassified by the model.
-\begin{figure}[tbp]
+\begin{figure}[p]
   \centering
   \includegraphics[width=\textwidth]{cpGraph_AllVars.png}
-  \caption{Cross validationerror versus the complexity parameter graph for the model developed with all variables and all individuals included.}
+  \caption{Cross validation error versus the complexity parameter graph for the model developed with all variables and all individuals included.}
   \label{fig:cpGraph_AllVars}
@@
 After producing this model and investigating figure~\ref*{cpGraph_AllVars},the cross validation error versus the complexity parameter graph, it was determined that no pruning was needed.
-No pruning was needed because as shown in figure~\ref*{cpGraph_AllVars} all of the different trees fall below the horizontal line plotted on the graph.
-Large error rate motivated data modification and reassessment.
+No pruning was needed because as shown in figure~\ref*{cpGraph_AllVars} all of the different trees calculated fall above the horizontal line plotted on the graph except for the tree with 7 terminal nodes which was the largest model created.
+The large error rate produced in the model with all individuals and variables motivated data modification and reassessment.
 \par
 The next model that was made trimmed the number of observations by if an individual had a serum cholesterol level of 0 mg/dL. 
-\begin{figure}[tbp]
+\begin{figure}[p]
   \centering
@@
 \end{figure}
-As shown in figure~\ref*{fig:cpGraph_NoCholesterol} it is seen that a tree with 2 terminal nodes,4 terminal nodes, 6 terminal nodes, 7 terminal nodes, and 11 terminal nodes all fit within the cirterion of having a relative error that is below the horizontal line plotted on the graph.
-Generally a smaller tree is reckognized as better because the computer has to do less work.
+As shown in figure~\ref*{fig:cpGraph_NoCholesterol} it is seen that a tree with 2 terminal nodes,4 terminal nodes, 6 terminal nodes, 7 terminal nodes, and 11 terminal nodes all fit within the criterion of having a relative error that is below the horizontal line plotted on the graph.
+Generally a smaller tree is recognized as better because the computer has to do less work.
 Therefore the two sincere contenders for this model are the tree with 2 terminal nodes and one split or the tree with 4 terminal nodes and 3 splits.
@@
 Additionally ST slope was also removed because another variable, Oldpeak, in the data set provides a numeric measurement for this.
-Also the ST Slope variable, similar to the chest pain type variable, has 3 different classifications instead of 2.
-\begin{figure}[tbp]
+The ST Slope variable, similar to the chest pain type variable, has 3 different classifications instead of 2.
+\begin{figure}[p]
   \centering
@@
 \end{figure} 
-This tree also showed qualities which would make it a good candidate for pruning.
+The tree with both the ST Slope and chest pain type variable removed also showed qualities which would make it a good candidate for pruning.
 As seen in figure~\ref*{fig:cpGraph_NoSTSandCPT} it is shown that both a tree with 3 terminal nodes and 2 splits or 6 terminal nodes and 5 splits would be optimal.
@@
 This means 49.24\% of the data in the validation set was misclassified.
-This model was surprisingly the one with the lowest error rate.
+The model, which removed the St Slope and chest pain type variable, was surprisingly the one with the lowest error rate.
 \par
-\begin{figure}[tbp]
+\begin{figure}[p]
   \centering
@@
 \end{figure} 
-The final and best decision tree out of the three models constructed in this paer is shown in figure~\ref*{fig:DTree_noSTSandCPT}.
+The final and best decision tree out of the three models constructed in this paper is shown in figure~\ref*{fig:DTree_noSTSandCPT}.
 Exercise Angina, Cholesterol, and Old Peak variables are deemed the most significant at determining heart disease in individuals in this model.
+Additionally as shown in figure~\ref*{fig:DTree_noSTSandCPT} exercise angina inhabit the first split, the second split is occupied by if an individual has a cholesterol value greater than of equal to 129 mg/dL, the third split is if an individual's ST slope wave depression is less than .75.
+The last two splits are first if the max heart rate of an individual is greater than or equal to 133 bpm and the last split is the old peak variable again however the classification is if an individual's ST slope is less than 2.5.
 \par
+\newpage
 \section{Discussion}
@@
 Since the first error rate was so large in the first model with all of the variables I was led to try different models to see if they would improve the error rate.
-While each error rate is not identical I would argue that the differences between them are insignificant.
+While each error rate for each model is not identical I would argue that the differences between them are insignificant.
 I wanted to create a model without the cholesterol values of 0 mg/dL because if the values reported are truly the serum levels of cholesterol it would be impossible if not extremely unhealthy for these to be 0 mg/dL.
@@
 I found that these models work best with variables that are either numeric where a cut off point can be determined or strictly binary variables.
-This led me to take out the chest pain type because it had 4 buckets instead of 2.
-I also took out the ST slope variable because it seemed redundant and it also had 3 buckets of classification instead of 2.
-This model technically ended up having the smallest error rate but it still was around 50\%.
+This led me to take out the chest pain type because it had 4 buckets of classification instead of 2.
+I also took out the ST slope variable because it seemed redundant because the OldPeak variable covers the same information except instead of up, down, and flat the OldPeak variable provides actual numbers.
+This model technically ended up having the smallest error rate but it still was very close to 50\%.
 \par
-Limitations in this paper mainly stem from the data source itself.
-Since there are over 100 individuals with claimed blood serum cholesterol levels of 0 mg/dL when this is not medically possible I think that data point may have been misrepresented.
+Limitations in this paper mainly stem from the data source itself and my personal knowledge.
+Since there are over 100 individuals with claimed blood serum cholesterol levels of 0 mg/dL when this is not medically possible I think that data points may have been misrepresented.
 Additionally even though there were a significant amount of variables provided I think the addition of more variables like weight, BMI, exercise frequency, or a quantitative measure of healthy lifestyle might make better less error prone models.
 I also think the method chosen to create the models was limiting.
-Since all of the models have a 50\% error rate I can conclude and reconfirm that classical decision trees are not be the best machine learning model for this data.
+Since all of the models have a 50\% error rate I can conclude and reconfirm that classical decision trees are not the best machine learning model for this data.
 At the time of writing the proposal for this project and then when developing the rough draft and doing the actual coding and model creation for this project I did not have the knowledge of many other methods of machine learning for binary variables at the time.
-As the semester has continued and my knowledge from other classes has expanded the breadth of statistical processes that I know, it would be incredibly beneficial to delve into models such as logistic regression, linear discriminant analysis, quadratic discriminant analysis, and deep learning models such as feed forward networks.
+As the semester has continued and my knowledge from other classes has expanded the breadth of statistical processes that I know, in the future it would be incredibly beneficial to delve into models such as logistic regression, linear discriminant analysis, quadratic discriminant analysis, and deep learning models such as feed forward networks.
 Finally while this data set was large and quite comprehensive with more time, money, and resources more individuals and variables could be added to the data set.
@@
 \bibliography{refs}
-\bibliographystyle{ama}
+\bibliographystyle{chicago}
 

### 2022-12-09 Code/varexp.R
added code for varible exploration

@@ -1,1 +1,37 @@
+#creating factors of all variables that are some form of binary
+Heart$HeartDisease<-factor(Heart$HeartDisease, levels=c(0,1), labels=c("no","yes"))
+Heart$Sex<- factor(Heart$Sex,) #female=1 male=2
+Heart$ChestPainType<- factor(Heart$ChestPainType, levels=c("TA", "ATA", "NAP","ASY"))
+#TA=1, ATA=2,NAP=3, ASY=4
+Heart$FastingBS<-factor(Heart$FastingBS, levels=c(0,1), labels=c("Fasting BS<120","Fasting BS>120"))
+Heart$RestingECG<- factor(Heart$RestingECG, levels=c("Normal", "ST"), labels=c("Normal","ST"))
+Heart$ExerciseAngina<-factor(Heart$ExerciseAngina, levels=c("N","Y"), labels=c("no","yes"))
+Heart$ST_Slope<- factor(Heart$ST_Slope, levels=c("Up", "Flat", "Down"), labels=c("Up","Flat","Down"))
 
+#variable exploration
+Heartnc<-subset(Heart,Heart$Cholesterol!=0 & Heart$RestingBP!=0)
+
+#sex
+counts<- table(Heart$Sex, Heart$HeartDisease)
+barplot(counts, main="Heart Disease and Sex", xlab="Heart Disease", ylab="Frequency", col=c("darksalmon","darkseagreen"), legend=rownames(counts))
+#age
+boxplot(Heart$Age~Heart$HeartDisease, main="Age and Heart Disease",xlab="Heart Disease", ylab="Age",col="cyan4" )
+#cholesterol 
+boxplot(Heartnc$Cholesterol~Heartnc$HeartDisease, main="Cholesterol and Heart Disease", xlab="Heart Disease", ylab="Serum Cholesterol (mg/dL) ", col="cyan4")
+#resting BP
+boxplot(Heart$RestingBP~Heart$HeartDisease, main="Resting Blood Pressure and Heart Disease", xlab="Heart Disease", ylab="Resting Blood Pressure (mmHg)", col="cyan4")
+#fasting BS need to figure out how to move legend
+counts2<- table(Heart$FastingBS, Heart$HeartDisease)
+barplot(counts2,main="Fasting Blood Sugar and Heart Disease", xlab="Heart Disease", ylab= "Frequency", col=c("cyan4","darkblue" ),legend=rownames(counts2))
+#resting ecg
+counts3<- table(Heart$RestingECG,Heart$HeartDisease)
[…]

### 2022-12-09 Code/no_STSlope_and_chestPainType_model.R
added code into file

@@ -1,1 +1,23 @@
-
+#retry with no chest pain type or ST slope
+set.seed(4321)
+train<-sample(918,459)
+Htrain<-Heart[train,]
+Htest<-Heart[-train,]
+#above is redundant code and same as the all variables and all individuals model
+library(rpart)
+set.seed(4321)
+ctree2<-rpart(HeartDisease ~ Age+Sex+RestingBP+Cholesterol+FastingBS+RestingECG+MaxHR+ExerciseAngina+Oldpeak, data=Htrain, method="class", parms=list(split="information"))
+library(rpart.plot)
+show(rpart.plot(ctree2))
+summary(ctree2)
+ctree2$cptable
+plotcp(ctree2)
+h2.predict<-predict(ctree2, data=Htest,type="class")
+h2.performance<- table(Htest$HeartDisease,h2.predict,dnn=c("Actual", "Predicted") )
+h2.performance
+ctree2.pruned<-prune(ctree2,cp=0.02941176)
+show(rpart.plot(ctree2.pruned))
+h3.predict<-predict(ctree2.pruned, data=Htest,type="class")
+h3.performance<- table(Htest$HeartDisease,h3.predict,dnn=c("Actual", "Predicted") )
+h3.performance
+#error rate= 226/459=.4924

### 2022-12-09 Code/no_cholesterol0_individuals_model.R
added code into the file

@@ -1,1 +1,24 @@
-
+Heartnc<-subset(Heart,Heart$Cholesterol!=0 & Heart$RestingBP!=0)
+set.seed(4321)
+train1<-sample(746,373)
+Hearttrain<-Heartnc[train1,]
+Hearttest<-Heartnc[-train1,]
+#creating the classical decision tree
+library(rpart)
+set.seed(4321)
+classtree<-rpart(HeartDisease ~ ., data=Hearttrain, method="class", parms=list(split="information"))
+library(rpart.plot)
+show(rpart.plot(classtree))
+summary(classtree)
+classtree$cptable
+plotcp(classtree)
+#the tree DOES need to be pruned
+#I think  option 4 (3 splits) is good
+classtree4.pruned<-prune(classtree, cp=.01436782)
+show(rpart.plot(classtree4.pruned))
+Heart.predict<-predict(classtree4.pruned,data=Hearttest, type="class" )
+Heart.performance<- table(Hearttest$HeartDisease,Heart.predict,dnn=c("Actual", "Predicted") )
+Heart.performance
+print(classtree)
+summary(classtree)
+# error rate= 194/373=.5201

### 2022-12-09 Code/All_vars_all_individuals_model.R
added code into the file

@@ -1,1 +1,19 @@
-
+#creating the test and training sets
+set.seed(4321)
+train<-sample(918,459)
+Htrain<-Heart[train,]
+Htest<-Heart[-train,]
+#creating classical decision tree (DTreeAllVars)
+library(rpart)
+set.seed(4321)
+ctree<-classtree<-rpart(HeartDisease ~ ., data=Htrain, method="class", parms=list(split="information"))
+library(rpart.plot)
+show(rpart.plot(ctree))
+summary(ctree)
+ctree$cptable
+plotcp(ctree)
+h.predict<-predict(ctree, data=Htest,type="class")
+h.performance<- table(Htest$HeartDisease,h.predict,dnn=c("Actual", "Predicted") )
+h.performance
+#the tree doesn't need to be pruned
+#error rate of 228/459=.496732 - note this fraction is number wrong/ total
